The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2013-11-08 sections/climate.tex
General footnotes clear-up

The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec})\qiong{Change unit to mm per month.}, primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades. A less pronounced precipitation peak is also visible over the interior ranges. From an ice sheet mass balance point of view, these heterogeneities become even stronger due to the difference in timing of the precipitation peak through the year. Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).

>>>>

The spatial distribution of JJA and DJF mean precipitation rates in WorldClim reveal a strong precipitation decline beyond the coastal regions (Fig.~\ref{fig:prec}), primarily as a result of the continuous orographic barrier formed by the Boundary Ranges, the Coast Mountains and the Cascades. A less pronounced precipitation peak is also visible over the interior ranges. From an ice sheet mass balance point of view, these heterogeneities become even stronger due to the difference in timing of the precipitation peak through the year. Whereas coastal regions experience most precipitation as snow during the accumulation season (DJF), inland regions experience dry winters and most of the annual precipitation consequently falls as rain during the ablation season (JJA).

### 2013-11-08 sections/discuss.tex
General footnotes clear-up

Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Fig.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we bound our analysis to the  WorldClim, ERA-Interim and NARR input climatologies. In order to compare modelled ice sheets of a similar size, we select for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$\irina{justify}. This corresponds to temperature offset values of 8 (WorldClim), 6 (ERA-Interim) and 7\,\unit{\degree C} (NARR). These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.

@@

Secondly, our simulations produce anomalous\irina{Any evidence that there was no ice cover there? Refer to the source.} ice cover on parts of the continental shelf in the Arctic Ocean. These regions experience a marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}). However in our simulations, sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land. This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.

@@

Our choice of 10\,ka simulation length could certainly be discussed further. However, our representation of climate history by constant temperature offsets is voluntarily simplistic. To better understand the transient character of the LGM Cordilleran Ice Sheet, time-dependent palaeo-climate forcing would be needed. This is left for future research.

@@

Our simplistic temperature offset method is certainly a crude simplification\irina{repetitive} of past climate changes over the region. Although it is clear that these temperature changes were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets. A more correct, yet more complex approach would be to use coupling to a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.

>>>>

Considering previously identified weaknesses of the CFSR and NCEP/NCAR data (Fig.~\ref{fig:tempheatmap}--\ref{fig:biatmbars}), we bound our analysis to the  WorldClim, ERA-Interim and NARR input climatologies. In order to compare modelled ice sheets of a similar size, we select for each climate forcing the simulation that leads to a final glaciated area closest to the approximate size of the LGM Cordilleran Ice Sheet of $2\,\times10^6\,\unit{km^2}$. This corresponds to temperature offset values of 8 (WorldClim), 6 (ERA-Interim) and 7\,\unit{\degree C} (NARR). These qualitative “best” runs are presented in Figure~\ref{fig:best} along with their associated temperature depressions and a reconstruction of the LGM ice sheet margin by \citet{dyke-2004}.

@@

Secondly, our simulations produce anomalous ice cover on parts of the continental shelf in the Arctic Ocean (Fig.~\ref{fig:best}). These regions experience a marine climate, including lower summer temperature than on the adjacent land (Fig.~\ref{fig:temp}). However in our simulations, sea-level is lowered by 120~m, turning large parts of the low-sloping continental shelf into land. This anomalous ice cover is to be interpreted as an artefact arising from our simplistic temperature offset method, and has little effect on the rest of the results.

@@

@@

Our choice of 10\,ka simulation length could certainly be discussed further. However, our representation of climate history by constant temperature offsets is voluntarily simplistic. To better understand the transient character of the LGM Cordilleran Ice Sheet, time-dependent palaeo-climate forcing would be needed. Although it is clear that past temperature changes over the region were neither homogeneous nor constant in time, and probably associated with precipitation changes, their patterns are not trivial and may potentially display a strong inter-dependence with the evolution of the Cordilleran and Laurentide ice sheets. A more correct, yet more complex approach would be to use coupling to a GCM of intermediate complexity \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}.

### 2013-11-08 sections/intro.tex
General footnotes clear-up

Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project (PMIP) \citep{joussaume-taylor-1995} provide perhaps a more reasonable representation of past climate. However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This condition potentially exerts influence on subsequently modelled ice sheet geometries, and particularly glacier extent.
\arjen{Has this been shown? Add ref.}

@@

Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs fromo several climate datasets. This is not entirely a new approach as a few studies of this kind are presently available. \citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis. \citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheet model to climate forcing from different PMIP LGM simulations\qiong{Detail in a few lines the conclusions of theses studies}. In our study, to limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data. Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}. Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western Northern America.

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap})\irina{include all geographic names mentioned in the text}\arjen{elevation scale white-end can be confused with glaciers} covered an area that presently experiences strong regional variations in climate. From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical and geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}. While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

@@

%While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, ours encompasses most of the LGM ice extent and adjacent non-glaciated parts of northern Yukon Territory and interior Alaska (Fig.~\ref{fig:locmap}).
%Our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response following the large advances made in scientific computing and ice sheet modelling since then.

>>>>

Climatologies obtained from GCM palaeo-climate simulations such as produced within the Paleoclimate Modelling Intercomparison Project (PMIP) \citep{joussaume-taylor-1995} provide perhaps a more reasonable representation of past climate. However, as such climatologies are only available for specific periods of time, using them as climate forcing for an ice sheet model requires either an assumption of steady-state ice sheet response to climatic fluctuations \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that GCM palaeo-climate simulations themselves rely on an accurate description of the global surface topography and therefore include global ice sheet reconstructions such as ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This condition could potentially exert influence on subsequently modelled ice sheet geometries, and particularly glacier extent.

@@

Here, we propose to address some of the uncertainties concerning climate forcing of numerical glacier models by evaluating the responses of a numerical model, in terms of glacier extent, to inputs fromo several climate datasets. This is not entirely a new approach as a few studies of this kind are presently available. \citet{quiquet-etal-2012}, for example, assessed the sensitivity of a Greenland ice sheet model to various atmospheric forcing, including a regional parametrization by \citet{fausto-etal-2009}, output from several GCMs and an atmospheric reanalysis. \citet{rodgers-etal-2004} and \citet{charbit-etal-2007} tested the sensitivity of a model of the Northern Hemisphere ice sheet model to climate forcing from different PMIP LGM simulations. Although using different set-ups, all three studies demonstrate the very large sensitivity of ice sheet model to the choice of climate forcing data used. In our study, to limit the degrees of freedom in our model and obtain results independent of palaeo-ice sheet reconstructions such as ICE-4G, we use a simple temperature offset similar to the approaches by \citet{greve-etal-1999} and \citet{bintanja-etal-2005} and assess ice sheet model sensitivity to the choice of present-day climate data. Hence, rather than using GCM output, we force our model with climate reanalysis data, which builds on observational information through data assimilation \citep{bengtsson-etal-2007}. Furthermore, we focus our study regionally on the former Cordilleran Ice Sheet in western Northern America.

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. From a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant stratigraphical and geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to simulate ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} and world-wide \citep{yoshimori-etal-2001}. While these studies generally reproduce the magnitude of North American glaciation at LGM reasonably well, there exists a tendency in the simulations that are independent of ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and interior Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

@@

### 2013-11-08 sections/model.tex
General footnotes clear-up

A pseudo-plastic sliding law \citep{aschwanden-etal-2013} (supplement) relates the bed-parallel shear stress and the sliding velocity. The yield stress is modelled using the Mohr-Coulomb criterion. The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of current bed elevation, with lowest values occurring at low elevations,
\julien{phi is independent from basal water which is computed by the model (and sliding depends on both) so the 10 to 30 degree range is meant to reflect different bedrock/sediment surfaces rather than dry vs wet beds.}

@@

where $\phi$ is given in degrees and $z$ in meters above current sea level. This models weakening of the till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a resolution of 1\,arc-minute \citep{data:etopo1}.\julien{NaturalEarth data are vectors (coastline, rivers, etc.) used in Fig. 1, not model input.} Sea level is lowered by 120\,m and basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.

>>>>

A pseudo-plastic sliding law \citep{aschwanden-etal-2013} (supplement) relates the bed-parallel shear stress and the sliding velocity. The yield stress is modelled using the Mohr-Coulomb criterion. The till friction angle $\phi$ varies from 10 to 30\degree\ and is a function of current bed elevation, with lowest values occurring at low elevations,

@@

where $\phi$ is given in degrees and $z$ in meters above current sea level. This models weakening of the till associated with the presence of marine sediments \citep{martin-etal-2011,aschwanden-etal-2013}. Basal topography (Fig.~\ref{fig:topo}) is derived from the ETOPO1 combined topography and bathymetry dataset with a resolution of 1\,arc-minute \citep{data:etopo1}. Sea level is lowered by 120\,m and basal topography responds to ice load following a regional isostasy model that includes lithospheric flexure and mantle relaxation \citep{lingle-clark-1985}.
